Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-11-17 manuscript.tex
latest

Far less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment. In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2012}. For example, MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. The r/K scheme's original purpose was as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of adaptive trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.

Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $r=b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$. Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

@@

$K$-selection in the sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1962}, would be represented in our model by smaller individual territorial requirements. To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density $L$. The situation is more complicated if those genotypes differ in multiple traits, and when the differences in territorial requirements become large enough that territorial contests can occur on different scales. We leave these complications for future work.

>>>>

Far less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment. In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2012}. For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. The r/K scheme's original purpose was as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.

@@

Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of adaptive trait evolution under different regimes of stress and disturbance. As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme. In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}. Thus, while trade-offs may amplify specialization, they are not necessary for it. As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances. Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.

Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967} since $b$ is closely related to the maximal, low-density growth rate $r=b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''. However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, these correlations are not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$. Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.

@@

$K$-selection in the sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, would be represented in our model by smaller individual territorial requirements. To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density $L$. The situation is more complicated if those genotypes differ in multiple traits, and when the differences in territorial requirements become large enough that territorial contests can occur on different scales. We leave these complications for future work.
